Staged Compilation With Two-Level Type Theory

### 2022-02-21 paper/paper.tex
stuff

\newcommand{\Bool}{\msf{Bool}}
\newcommand{\true}{\msf{true}}
\newcommand{\false}{\msf{false}}
\newcommand{\List}{\msf{List}}
\newcommand{\nil}{\msf{nil}}
\newcommand{\cons}{\msf{cons}}
\newcommand{\Nat}{\msf{Nat}}
\newcommand{\zero}{\msf{zero}}
\newcommand{\suc}{\msf{suc}}
\renewcommand{\tt}{\msf{tt}}
\newcommand{\fst}{\msf{fst}}
\newcommand{\snd}{\msf{snd}}
\newcommand{\mylet}{\msf{let}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\id}{\msf{id}}

@@

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

@@

\begin{abstract}
  The aim of staged compilation is to enable metaprogramming in a way such that
  we have guarantees about the well-formedness of code output, and we can also
  mix together object-level and meta-level code in a concise and convenient
  manner. In this work, we observe that two-level type theory (2LTT), a system
  originally devised for the purpose of developing synthetic homotopy theory,
  also serves as a system for staged compilation with dependent types. 2LTT has
  numerous good properties for this use case: it has a concise specification,
  well-developed model theory, and it supports a wide range of language features
  both at the object and the meta level. First, we give an overview of 2LTT's
  features and applications in staging. Then, we present a staging algorithm and
  provide a proof of correctness. Our algorithm is ``staging-by-evaluation'',
  analogously to the technique of normalization-by-evaluation, in that staging
  is given by the evaluation of 2LTT syntax in a semantic domain. The staging
  algorithm together with its correctness proof constitutes a proof of strong
  conservativity of 2LLT over the object theory. To our knowledge, this is the
  first system for staged compilation which supports full dependent types and
  unrestricted staging for types.
\end{abstract}

@@

\subsection{Overview \& Contributions}

@@

\subsection{Staged Programming in 2LTT}

@@

\subsubsection{Monomorphization}

@@

\subsection{Models and Syntax of 2LTT}
We use an algebraic specification for models, and specify the syntax as the
initial model. We only handle well-formed syntactic objects and only define
operations which respect syntactic definitional equality; in fact, we identify
definitional equality with metatheoretic equality. First, we define the
structural scaffolding of 2LTT without type formers. The following specification
is agnostic about the sizes of $\Set$-s involved, i.e.\ we can model underlying
sorts using any metatheoretic $\Set_j$.

@@

\begin{definition} A \textbf{model of 2LTT} is a model of basic 2LTT
which supports certain type formers. For the sake of brevity, we concretely
specify here a small collection of type formers. However, we will argue that our
results extend to more general notions of inductive types. We specify type
formers in the following. We assume that all type formers are natural,
i.e.\ stable under substitution.
\begin{itemize}
\item \emph{Universes}. For each $i$ and $j$, we have a Coquand-style universe
  \cite{TODO} in $\Ty_{i,j}$. This consists of $\U_{i,j} : \Ty_{i,j+1}\,\Gamma$,
  together with $\El : \Tm_{i,j+1}\,\Gamma\,\U_{i,j} \to \Ty_{i,j}\,\Gamma$ and $\Code$, where $\Code$
  and $\El$ are inverses.
\item \emph{$\Sigma$-types}. We have $\Sigma\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$
  and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism of pairing and projections:
  \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,A\,B) \simeq ((t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
  We write $(t,\,u)$ for pairing and $\fst$ and $\snd$ for projections.
\item \emph{Function types}. We have $\Pi\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A :
  \Ty_{i,j}\,\Gamma$ and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with
  $\app : \Tm_{i,j}\,\Gamma\,(\Pi\,A\,B) \to \Tm_{i,j}\,(\Gamma \ext (x :
  A))\,B$ and its inverse $\lam$.
\item \emph{Natural numbers}. We have $\Nat_{i,j} : \Ty_{i,j}\,\Gamma$, $\zero_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$,
  and $\suc_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j} \to \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$. The eliminator is the following.
  \begin{alignat*}{5}
    & \NatElim :\hspace{0.5em}
                &&  (P &&: \Ty_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \\
    &           &&  (z &&: \Tm_{i,k}\,\Gamma\,(P[n\mapsto \zero_{i,j}])) \\
    &           &&  (s &&: \Tm_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \ext (\mit{pn} : P[n\mapsto n]))\,(P[n\mapsto \suc_{i,j}\,n]))\\
    &           &&  (t &&: \Tm_{i,j}\,\Gamma\,\Nat_{i,j}) \\
    &\hspace{3em}\to        &&&& \Tm_{i,j}\,\Gamma\,(P[n\mapsto t]))
  \end{alignat*}
  We also have the $\beta$-rules:
  \begin{alignat*}{5}
    & \NatElim\,P\,z\,s\,\zero_{i,j}     &&= z \\
    & \NatElim\,P\,z\,s\,(\suc_{i,j}\,t) &&= s[n \mapsto t,\,\mit{pn} \mapsto \NatElim\,P\,z\,s\,t]
  \end{alignat*}
  Note that we can eliminate from a level $j$ to any level $k$.
\end{itemize}
\end{definition}

\begin{definition}
The \textbf{syntax of 2LTT} is defined to be the initial model of 2LTT, where
all underlying sets are in $\Set_0$. The syntax can be assumed to exist, since
the notion of model is algebraic; the specification can be expressed using a
suitable notion of algebraic signatures, such as one for essentially algebraic
theories \cite{TODO} or as a finitary quotient inductive-inductive signature
\cite{TODO}. The initiality of syntax directly yields a notion of
\emph{recursion on syntax}, and it can be shown that initiality also implies an
\emph{induction principle} \cite{TODO}.
\end{definition}

The examples in Section \ref{TODO} can be faithfully represented in the formal
syntax. The only notable difference is that the formal syntax uses Coquand-style
universes instead of Russell-style ones. We use the former because it is much
easier to model in the semantics in Section \ref{TODO}. We note though that it
is straightforward to elaborate from a Russell-style surface syntax to the
formal version, by inserting $\El$ and $\Code$ as required.

@@

\begin{definition}
We have an \textbf{embedding} of object syntax syntax into 2LTT syntax, consisting
of the following functions:
\begin{alignat*}{4}
  & \emb{\blank} : \Con\ob \to \Con\\
  & \emb{\blank} : \Sub\ob\,\Gamma\,\Delta \to \Sub\,\emb{\Gamma}\,\emb{\Delta}\\
  & \emb{\blank} : \Ty_{\mbbo j}\,\Gamma \to \Ty_{0,j}\,\emb{\Gamma}\\
  & \emb{\blank} : \Tm_{\mbbo j}\,\Gamma\,A \to \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}
\end{alignat*}
Embedding maps all type and term formers to the corresponding ones in 2LTT, and
strictly preserves all structure.
\end{definition}

@@

\begin{definition}
A \textbf{staging algorithm} consists of two functions:
  \begin{alignat*}{3}
    & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma \\
    & \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A)
  \end{alignat*}
Note that we can stage open types and terms as long as their contexts are purely
object-level. By \emph{closed staging} we mean staging only for closed types and
terms.
\end{definition}
\begin{definition}
  A staging algorithm $\Stage$ is \textbf{correct} if the following properties hold:
  \begin{itemize}
  \item \emph{Soundness:} $\emb{\Stage\,A} = A$ and $\emb{\Stage\,t} = t$.
  \item \emph{Completeness:} staging respects definitional equality.
  \item \emph{Stability:} $\Stage\,\emb{A} = A$ and $\Stage\,\emb{t} = t$.
  \end{itemize}
\end{definition}
We make some remarks on correctness. First, we get completeness for free in our
algebraic setup, since \emph{all} functions definable in the metatheory must
respect definitional equality.

Second, note that soundness and stability together is the statement that
embedding is a bijection on types and terms. This is a statement of
\emph{conservativity} of 2LTT over the object theory. In \cite{TODO} a
significantly weaker conservativity theorem is shown, which expresses that there
exists a function from $\Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$ to $\Tm_{\mbbo
  j}\,\Gamma\,A$.

Lastly, we note that the correctness terminology is borrowed from previous works
on normalization-by-evaluation, where the soundness-completeness-stability trio
has been used several times \cite{TODO}. In \cite{cubicalnbe}, stability is
instead called ``idempotence''.

@@

We review the presheaf model described in \cite[Section~2.5.3]{twolevel}.  This
will directly yield a closed staging algorithm, by evaluation of 2LTT types and
terms in the model. Here, contexts are presheaves over the syntactic category of
the object theory. We denote the components of the model by putting hats on 2LTT
components, e.g.\ as in $\hCon$. We summarize the key components in the
following.

@@

\begin{definition}
$\bs{\hTy_{1,j}\,\Gamma : \Set_\omega}$ is the set of \emph{displayed presheaves}
over $\Gamma$; see e.g.\ \cite{TODO}. This is equivalent to the set of
presheaves over the category of elements of $\Gamma$, but it is usually more
convenient in calculations. An $A : \hTy\,\Gamma$ has an action on objects
$|A| : \{a : \Cono\} \to |\Gamma|\,a \to \Set_j$ and an action on
morphisms $\blank[\blank] : |A|\,\gamma \to (f : \Subo\,a\,b) \to
|A|\,(\gamma[f])$, such that $\alpha[\id] = \alpha$ and $\alpha[f \circ g] = \alpha[f][g]$.

@@

%% \begin{definition}
%% $\bs{\wh{\U}_{1,j} : \hTy_{1,j+1}\,\Gamma}$ is defined by
%% $|\wh{\U}_{1,j}|\,\{a\}\,\_ = \hTy_{1,j}\,(\y\,a)$, and defining action on
%% morphisms by $\hTy$ substitution. Note that the universe levels check out in
%% this definition, since by unfolding $\hTy_{1,j}\,(\y\,a)$ we can show that it
%% is in $\Set_j$.
%% \end{definition}

@@

\begin{definition}
$\bs{{\widehat{\emptycon}} : \hCon}$ is defined as the presheaf
which is constantly $\top$, i.e.\ $|\wh{\emptycon}|\,\_ = \top$.
\end{definition}

\begin{definition} For $A : \hTy_{1,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
pointwise by $|\Gamma\,\,\wh{\ext}\,A|\,a := (\gamma : |\Gamma|\,a) \times |A|\,\gamma$
and $(\gamma,\,\alpha)[f] := (\gamma[f],\,\alpha[f])$.
\end{definition}

Using the above definitions, we can model the syntactic category of 2LTT, and
also the meta-level family structure and all meta-level type formers. For an
exposition in previous literature, see \cite{TODO} and \cite{TODO}.

@@

Using the above definitions and following \cite{TODO}, we can model all type
formers in $\hTy_{0,j}$. Intuitively, that is because $\hTy_{0,j}$ and
$\hTm_{0,j}$ return types and terms, so we can reuse the type and term formers
in the object theory.

@@

\begin{definition}
$\bs{\wh{\Lift} A : \hTy_{1,j}\,\Gamma}$ is defined as $\Tm_{\mbbo j}[A]$. With
  this, we get that $\hTm_{0,j}\,\Gamma\,A$ is equal to
  $\hTm_{1,j}\,\Gamma\,(\wh{\Lift}\,A)$. Hence, we can define both quoting and
  splicing as identity functions in the model.
\end{definition}

@@

\begin{definition} The \textbf{evaluation morphism, denoted $\bs{\ev}$} is
a 2LTT model morphism from the syntax of 2LTT to $\hato$, arising from the initiality
of the syntax. In other words, $\ev$ is defined by recursion on the syntax and
strictly preserves all structure.
\end{definition}

@@

What about general (open) staging though? Given $A : \Ty_{0,j}\,\emb{\Gamma}$,
we get $|\ev\,A|\,\{\Gamma\} : |\emb{\Gamma}|\,\Gamma \to \Ty_{\mbbo j}$. We
need an element of $|\emb{\Gamma}|\,\Gamma$, in order to extract an object-level
type. Such ``generic'' semantic contexts should be possible to construct, because
elements of $|\emb{\Gamma}|\,\Gamma$ are essentially lists of object-level terms
in $\Gamma$, by Definition \ref{TODO}, so $|\emb{\Gamma}|\,\Gamma$ should be
isomorphic to $\Subo\,\Gamma\,\Gamma$.  It turns out that this falls out from
the stability proof of $\ev$.

@@

\begin{definition} We define \textbf{open staging} as follows.
\begin{alignat*}{4}
  & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma \hspace{2em} && \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A) \\
  & \Stage\,A := |\ev\,A|\,\Gamma^P && \Stage\,t := |\ev\,t|\,\Gamma^P
\end{alignat*}
\end{definition}

@@

We have defined staging in a constructive metatheory, so it is possible to
extract an implementation from it. We discuss its operational behavior and
potential optimizations in the following. First, consider the computational
part of evaluation:
\begin{alignat*}{3}
  & |\blank| \circ \ev : \Ty_{1 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Set_j\\
  & |\blank| \circ \ev : \Ty_{0 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Ty_{\mbbo j}\,\Delta\\
  & |\blank| \circ \ev : \Tm_{1 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
    |\ev\,A|\,\gamma \\
  & |\blank| \circ \ev : \Tm_{0 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
    \Tm_{\mbbo\,j}\,\Delta\,(|\ev\,A|\,\gamma)
\end{alignat*}
Operationally, we interpret syntactic types or terms in a semantic environment
$\gamma : |\ev\,\Gamma|\,\Delta$. These environments are lists containing
object-level terms and semantic values. The semantic values are represented
using metatheoretic inductive types and function types.
\begin{itemize}
  \item $|\Nat_1|\,\_$ is simply the set of natural numbers $\mbb{N}$.
  \item $|\Sigma_1\,A\,B|\,\gamma$ is simply a set of pairs of values.
  \item In the non-dependent case, a function type $A\,\wh{\to}\,B$ is defined as the presheaf
        exponential \cite{TODO}. The computational part of $|A\,\wh{\to}\,B|\,\{\Delta\}\,\gamma$
        is given by a function with type
        \[ (\Theta : \Cono) \to (\sigma : \Subo\,\Theta\,\Delta) \to |A|\,(\gamma[\sigma]) \to |B|\,(\gamma[\sigma])        \]
        This may be also familiar as the semantic implication from the Kripke
        semantics of intuitionistic logics. Whenever we evaluate a function
        application, we supply an extra $\id : \Subo\,\Delta\,\Delta$. This may
        incur cost via the $\gamma[\id]$ restrictions in a naive
        implementation, but this is easy to optimize, by introducing a formal
        representation of $\id$, such that $\gamma[\id]$ immediately computes to
        $\gamma$. The dependent case is analogous operationally.
\end{itemize}
In summary, in the meta-level fragment of 2LTT, $\ev$ yields a reasonably efficient
computation of closed values, which reuses functions from the ambient
metatheory. Alternatively, instead of using ambient functions, we could use our
own implementation of function closures during staging.

@@

The exact same efficiency issue occurs in formal presheaf-based
normalization-by-evaluation \cite{TODO}. However, in practical implementations
\cite{TODO} this issue can be fully solved by using De Bruijn \emph{levels} in
the semantic domain, thus arranging that weakening has no operational cost on
semantic values. This corresponds to Abel's ``liftable terms'' presentation
\cite{TODO}. We can use the same solution in staging, by using de Bruijn levels
and delayed weakenings in object-level terms during staging. It is a somewhat
unfortunate mismatch that indices are far more convenient in formalization, but
levels are more efficient in practice.

In our prototype implementation, we use the above optimization with De Bruijn
levels, and we also switch to \emph{syntax-directed} staging. In the formal
staging algorithm, since types are tracked, quotes and splices have no
operational significance; recall that they are interpreted as identity functions
in Definition \ref{TODO}. However, it seems more efficient to not compute types
at all and instead rely on quotes and splices to keep track of the current
stage.

>>>>

\newcommand{\Bool}{\msf{Bool}}
\newcommand{\true}{\msf{true}}
\newcommand{\false}{\msf{false}}
\newcommand{\True}{\msf{True}}
\newcommand{\False}{\msf{False}}
\newcommand{\List}{\msf{List}}
\newcommand{\nil}{\msf{nil}}
\newcommand{\cons}{\msf{cons}}
\newcommand{\Nat}{\msf{Nat}}
\newcommand{\zero}{\msf{zero}}
\newcommand{\suc}{\msf{suc}}
\renewcommand{\tt}{\msf{tt}}
\newcommand{\fst}{\msf{fst}}
\newcommand{\snd}{\msf{snd}}
\newcommand{\mylet}{\msf{let}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\id}{\msf{id}}

@@

%%
%% end of the preamble, start of the body of the document source.
\hypersetup{draft}
\begin{document}

@@

\begin{abstract}
  The aim of staged compilation is to enable metaprogramming in a way such that
  we have guarantees about the well-formedness of code output, and we can also
  mix together object-level and meta-level code in a concise and convenient
  manner. In this work, we observe that two-level type theory (2LTT), a system
  originally devised for the purpose of developing synthetic homotopy theory,
  also serves as a system for staged compilation with dependent types. 2LTT has
  numerous good properties for this use case: it has a concise specification,
  well-behaved model theory, and it supports a wide range of language features
  both at the object and the meta level. First, we give an overview of 2LTT's
  features and applications in staging. Then, we present a staging algorithm and
  provide a proof of correctness. Our algorithm is ``staging-by-evaluation'',
  analogously to the technique of normalization-by-evaluation, in that staging
  is given by the evaluation of 2LTT syntax in a semantic domain. The staging
  algorithm together with its correctness proof constitutes a proof of strong
  conservativity of 2LLT over the object theory. To our knowledge, this is the
  first description of staged compilation which supports full dependent types and
  unrestricted staging for types.
\end{abstract}

@@

\subsection{Overview \& Contributions}\label{sec:overview}

@@

\subsection{Staged Programming in 2LTT}\label{sec:staged-programming}

@@

\subsubsection{Monomorphization}\label{sec:monomorphization}

@@

\subsection{Models and Syntax of 2LTT}
We use an algebraic specification for models, and specify the syntax as the
initial model. We only handle well-formed syntactic objects and only define
operations which respect syntactic definitional equality; in fact, we identify
definitional equality with metatheoretic equality. First, we define the
structural scaffolding of 2LTT without type formers, mostly following
\cite{twolevel}. The specification is agnostic about the sizes of sets involved,
i.e.\ we can model underlying sorts using any metatheoretic $\Set_j$.

@@

\begin{definition} A \textbf{model of 2LTT} is a model of basic 2LTT
which supports certain type formers. For the sake of brevity, we concretely
specify here a small collection of type formers. However, we will argue that our
results extend to more general notions of inductive types. We specify type
formers in the following. We assume that all type formers are natural,
i.e.\ stable under substitution.
\begin{itemize}
\item \emph{Universes}. For each $i$ and $j$, we have a Coquand-style universe
  \cite{coquand2018canonicity} in $\Ty_{i,j}$. This consists of $\U_{i,j} : \Ty_{i,j+1}\,\Gamma$,
  together with $\El : \Tm_{i,j+1}\,\Gamma\,\U_{i,j} \to \Ty_{i,j}\,\Gamma$ and $\Code$, where $\Code$
  and $\El$ are inverses.
\item \emph{$\Sigma$-types}. We have $\Sigma\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$
  and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism of pairing and projections:
  \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,A\,B) \simeq ((t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
  We write $(t,\,u)$ for pairing and $\fst$ and $\snd$ for projections.
\item \emph{Function types}. We have $\Pi\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A :
  \Ty_{i,j}\,\Gamma$ and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with
  $\app : \Tm_{i,j}\,\Gamma\,(\Pi\,A\,B) \to \Tm_{i,j}\,(\Gamma \ext (x :
  A))\,B$ and its inverse $\lam$.
\item \emph{Natural numbers}. We have $\Nat_{i,j} : \Ty_{i,j}\,\Gamma$, $\zero_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$,
  and $\suc_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j} \to \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$. The eliminator is the following.
  \begin{alignat*}{5}
    & \NatElim :\hspace{0.5em}
                &&  (P &&: \Ty_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \\
    &           &&  (z &&: \Tm_{i,k}\,\Gamma\,(P[n\mapsto \zero_{i,j}])) \\
    &           &&  (s &&: \Tm_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \ext (\mit{pn} : P[n\mapsto n]))\,(P[n\mapsto \suc_{i,j}\,n]))\\
    &           &&  (t &&: \Tm_{i,j}\,\Gamma\,\Nat_{i,j}) \\
    &\hspace{3em}\to        &&&& \Tm_{i,j}\,\Gamma\,(P[n\mapsto t]))
  \end{alignat*}
  We also have the $\beta$-rules:
  \begin{alignat*}{5}
    & \NatElim\,P\,z\,s\,\zero_{i,j}     &&= z \\
    & \NatElim\,P\,z\,s\,(\suc_{i,j}\,t) &&= s[n \mapsto t,\,\mit{pn} \mapsto \NatElim\,P\,z\,s\,t]
  \end{alignat*}
  Note that we can eliminate from a level $j$ to any level $k$.
\end{itemize}
\end{definition}

\begin{definition}
The \textbf{syntax of 2LTT} is defined to be the initial model of 2LTT, where
all underlying sets are in $\Set_0$. The syntax can be assumed to exist, since
the notion of model is algebraic; the specification can be expressed using a
suitable notion of algebraic signatures, such as one for essentially algebraic
theories \cite{partialhorn} or as a finitary quotient inductive-inductive
signature \cite{kaposi2019constructing}. The initiality of syntax directly
yields a notion of \emph{recursion on syntax}, and it can be shown that
initiality also implies an \emph{induction principle}.
\end{definition}

The examples in Section \ref{sec:tour-of-2ltt} can be faithfully represented in
the formal syntax. The only notable difference is that the formal syntax uses
Coquand-style universes instead of Russell-style ones. We use the former because
it is much easier to model in the semantics in Section \ref{sec:staging-algorithm}. We note
though that it is straightforward to elaborate from a Russell-style surface
syntax to the formal version, by inserting $\El$ and $\Code$ as required.

@@

\begin{definition}
We have an \textbf{embedding} of object syntax syntax into 2LTT syntax, consisting
of the following functions:
\begin{alignat*}{4}
  & \emb{\blank} : \Con\ob \to \Con\\
  & \emb{\blank} : \Sub\ob\,\Gamma\,\Delta \to \Sub\,\emb{\Gamma}\,\emb{\Delta}\\
  & \emb{\blank} : \Ty_{\mbbo j}\,\Gamma \to \Ty_{0,j}\,\emb{\Gamma}\\
  & \emb{\blank} : \Tm_{\mbbo j}\,\Gamma\,A \to \Tm_{0,j}\,\emb{\Gamma}\,\emb{A}
\end{alignat*}
Embedding strictly preserves all structure.
\end{definition}

@@

\begin{definition}\label{def:staging}
A \textbf{staging algorithm} consists of two functions:
  \begin{alignat*}{3}
    & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma \\
    & \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A)
  \end{alignat*}
Note that we can stage open types and terms as long as their contexts are purely
object-level. By \emph{closed staging} we mean staging only for closed types and
terms.
\end{definition}
\begin{definition}
  A staging algorithm $\Stage$ is \textbf{correct} if the following properties hold:
  \begin{enumerate}
  \item \emph{Soundness:} $\emb{\Stage\,A} = A$ and $\emb{\Stage\,t} = t$.
  \item \emph{Completeness:} staging respects definitional equality.
  \item \emph{Stability:} $\Stage\,\emb{A} = A$ and $\Stage\,\emb{t} = t$.
  \end{enumerate}
\end{definition}
We make some remarks on correctness. First, we get completeness for free in our
algebraic setup, since \emph{all} functions definable in the metatheory must
respect definitional equality.

Second, note that soundness and stability together is the statement that
embedding is a bijection on types and terms. This is a statement of
\emph{conservativity} of 2LTT over the object theory. In \cite{twolevel} a
significantly weaker conservativity theorem is shown, which expresses that there
exists a function from $\Tm_{0,j}\,\emb{\Gamma}\,\emb{A}$ to $\Tm_{\mbbo
  j}\,\Gamma\,A$.

Lastly, we note that the correctness terminology is borrowed from previous works
on normalization-by-evaluation, where the soundness-completeness-stability trio
has been used \cite{altenkirch1995categorical,altenkirch2016normalisation}. In
\cite{cubicalnbe}, stability is instead called ``idempotence''.

@@

We review the presheaf model described in \cite[Section~2.5.3]{twolevel}.  This
will directly yield a closed staging algorithm, by evaluation of 2LTT types and
terms in the model. First, we give a higher-level overview of this semantics.

TODO explanation here.
%% Here, contexts are presheaves over the syntactic category of
%% the object theory.

We proceed to summarize the key components of the model in a bit more detail. We
denote the components of the model by putting hats on 2LTT components, e.g.\ as
in $\hCon$. We summarize the key components in the following.

@@

\begin{definition}
$\bs{\hTy_{1,j}\,\Gamma : \Set_\omega}$ is the set of \emph{displayed
presheaves} over $\Gamma$; see e.g.\ \cite[Chapter~1.2]{huber-thesis}. This is
equivalent to the set of presheaves over the category of elements of $\Gamma$,
but it is usually more convenient in calculations. An $A : \hTy\,\Gamma$ has
an action on objects $|A| : \{a : \Cono\} \to |\Gamma|\,a \to \Set_j$ and an
action on morphisms $\blank[\blank] : |A|\,\gamma \to (f : \Subo\,a\,b) \to
|A|\,(\gamma[f])$, such that $\alpha[\id] = \alpha$ and $\alpha[f \circ g] =
\alpha[f][g]$.

@@

@@

\begin{definition}\label{def:psh-emptycon}
$\bs{{\widehat{\emptycon}} : \hCon}$ is defined as the presheaf
which is constantly $\top$, i.e.\ $|\wh{\emptycon}|\,\_ = \top$.
\end{definition}

\begin{definition}\label{def:psh-ext} For $A : \hTy_{1,j}\,\Gamma$, we define $\bs{\Gamma\,\,\wh{\ext}\,A}$
pointwise by $|\Gamma\,\,\wh{\ext}\,A|\,a := (\gamma : |\Gamma|\,a) \times |A|\,\gamma$
and $(\gamma,\,\alpha)[f] := (\gamma[f],\,\alpha[f])$.
\end{definition}

Using the above definitions, we can model the syntactic category of 2LTT, and
also the meta-level family structure and all meta-level type formers. For
expositions in previous literature, see \cite[Chapter~1.2]{huber-thesis} and
\cite[Section~4.1]{Hofmann97syntaxand}.

@@

Using the above definitions and following \cite{twolevel}, we can model all type
formers in $\hTy_{0,j}$. Intuitively, that is because $\hTy_{0,j}$ and
$\hTm_{0,j}$ return types and terms, so we can reuse the type and term formers
in the object theory.

@@

\begin{definition}\label{def:psh-lifting}
$\bs{\wh{\Lift} A : \hTy_{1,j}\,\Gamma}$ is defined as $\Tm_{\mbbo j}[A]$. With
  this, we get that $\hTm_{0,j}\,\Gamma\,A$ is equal to
  $\hTm_{1,j}\,\Gamma\,(\wh{\Lift}\,A)$. Hence, we can define both quoting and
  splicing as identity functions in the model.
\end{definition}

@@

\begin{definition} The \textbf{evaluation morphism, denoted $\bs{\ev}$}, is
a 2LTT model morphism from the syntax of 2LTT to $\hato$, arising from the initiality
of the syntax. In other words, $\ev$ is defined by recursion on the syntax and
strictly preserves all structure.
\end{definition}

@@

What about general (open) staging though? Given $A : \Ty_{0,j}\,\emb{\Gamma}$,
we get $|\ev\,A|\,\{\Gamma\} : |\emb{\Gamma}|\,\Gamma \to \Ty_{\mbbo j}$. We
need an element of $|\emb{\Gamma}|\,\Gamma$, in order to extract an object-level
type. Such ``generic'' semantic contexts should be possible to construct, because
elements of $|\emb{\Gamma}|\,\Gamma$ are essentially lists of object-level terms
in $\Gamma$, by Definitions \ref{def:psh-ext} and \ref{def:psh-emptycon}, so $|\emb{\Gamma}|\,\Gamma$ should be
isomorphic to $\Subo\,\Gamma\,\Gamma$.  It turns out that this falls out from
the stability proof of $\ev$.

@@

\begin{definition}\label{def:open-staging} We define \textbf{open staging} as follows.
\begin{alignat*}{4}
  & \Stage : \Ty_{0,j}\,\emb{\Gamma} \to \Ty_{\mbbo j}\,\Gamma \hspace{2em} && \Stage : \Tm_{0,j}\,\emb{\Gamma}\,A \to \Tm_{\mbbo j}\,\Gamma\,(\Stage\,A) \\
  & \Stage\,A := |\ev\,A|\,\Gamma^P && \Stage\,t := |\ev\,t|\,\Gamma^P
\end{alignat*}
\end{definition}

@@

We have defined staging in a constructive metatheory, so it is possible to
extract an implementation from it. We discuss its operational behavior and
potential optimizations in the following. First, consider the computational
part of evaluation:
\begin{alignat*}{3}
  & |\blank| \circ \ev : \Ty_{1 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Set_j\\
  & |\blank| \circ \ev : \Ty_{0 j}\,\Gamma \to \{\Delta : \Cono\} \to |\ev\,\Gamma|\,\Delta \to \Ty_{\mbbo j}\,\Delta\\
  & |\blank| \circ \ev : \Tm_{1 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
    |\ev\,A|\,\gamma \\
  & |\blank| \circ \ev : \Tm_{0 j}\,\Gamma\,A \to \{\Delta : \Cono\} \to (\gamma : |\ev\,\Gamma|\,\Delta) \to
    \Tm_{\mbbo\,j}\,\Delta\,(|\ev\,A|\,\gamma)
\end{alignat*}
Operationally, we interpret syntactic types or terms in a semantic environment
$\gamma : |\ev\,\Gamma|\,\Delta$. These environments are lists containing
object-level terms and semantic values. The semantic values are represented
using metatheoretic inductive types and function types.
\begin{itemize}
  \item $|\Nat_1|\,\_$ is simply the set of natural numbers $\mbb{N}$.
  \item $|\Sigma_1\,A\,B|\,\gamma$ is simply a set of pairs of values.
  \item In the non-dependent case, a function type $A\,\wh{\to}\,B$ is defined as the presheaf
        exponential. The computational part of $|A\,\wh{\to}\,B|\,\{\Delta\}\,\gamma$
        is given by a function with type
        \[ (\Theta : \Cono) \to (\sigma : \Subo\,\Theta\,\Delta) \to |A|\,(\gamma[\sigma]) \to |B|\,(\gamma[\sigma])        \]
        This may be also familiar as the semantic implication from the Kripke
        semantics of intuitionistic logics. Whenever we evaluate a function
        application, we supply an extra $\id : \Subo\,\Delta\,\Delta$. This may
        incur cost via the $\gamma[\id]$ restrictions in a naive
        implementation, but this is easy to optimize, by introducing a formal
        representation of $\id$, such that $\gamma[\id]$ immediately computes to
        $\gamma$. The dependent case is analogous operationally.
\end{itemize}
In summary, in the meta-level fragment of 2LTT, $\ev$ yields a reasonably efficient
computation of closed values, which reuses functions from the ambient
metatheory. Alternatively, instead of using ambient functions, we could use our
own implementation of function closures during staging.

@@

The exact same efficiency issue occurs in formal presheaf-based
normalization-by-evaluation, such as in \cite{kaposinbe} and
\cite{coquand2018canonicity}. However, in practical implementations this issue
can be fully solved by using De Bruijn \emph{levels} in the semantic domain,
thus arranging that weakening has no operational cost on semantic values; see
\cite{coquand1996algorithm}, \cite{modulartc} and \cite{untypedtc} for
implementations in this style. We can use the same solution in staging, by using
de Bruijn levels and delayed weakenings in object-level terms during staging. It
is a somewhat unfortunate mismatch that indices are far more convenient in
formalization, but levels are more efficient in practice.

In our prototype implementation, we use the above optimization with De Bruijn
levels, and we also switch to \emph{syntax-directed} staging. In the formal
staging algorithm, since types are tracked, quotes and splices have no
operational significance; recall that they are interpreted as identity functions
in Definition \ref{def:psh-lifting}. However, it seems more efficient to not compute types
at all and instead rely on quotes and splices to keep track of the current
stage.
